---
title: "Navigating MOSAIC"
author: "http://mosaicdatabase.web.ox.ac.uk"
date: "Updated 11 February 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vignette #1 - Navigating MOSAIC

Welcome to the **MOSAIC database**, a database of functional traits for comparative demography. The database, user, guide, and additional information can be found and <http://mosaicdatabase.web.ox.ac.uk>.

MOSAIC is a database that aggregates existing databases and adds new records for functional traits that currently do not have a database established. In this vignette, we will show you how to download the dataset, search records, and relate MOSAIC records with COMADRE, COMPADRE, and PADRINO databases.


##### <span style="color:cadetblue2">Optional clearance of working space</span>.
```{r eval=FALSE}
rm(list = ls()) # Clear your environment
if(!is.null(dev.list())) dev.off() # Clear plots/graphics
cat("\014") # Clear console

```

## Downloading MOSAIC

MOSAIC can be downloaded as an S4 object by running the below code in R. S4 data objects in R are an object oriented system in the R language that tightly regulate linked data objects and that works well for organising complex, hierarchical data. Similar to the more commonly used S3 objects (which use the "$" operator), S4 are comprised of objects that can be searched with the "@" operator, discussed in more detail below.

```{r echo=T, cache=FALSE, results=FALSE, warning=FALSE, comment=FALSE, warning=FALSE, message=FALSE}
library(devtools)
library(Rcompadre)
source_url("https://raw.githubusercontent.com/mosaicdatabase/mosaicdatabase/main/mosaic_fetch.R") # Link GitHub
mosaic <- mos_fetch("v1.0.0") # Loads version 1.0.0 of the mosaic database
rm(metaIndexer, mos_fetch, url, Indices) # Clean the working environment - removes fetch-related functions
```


## Basics of manually navigating MOSAIC



Unlike S3 objects, whose attributes/fields are searched with the "$" operator, S4 objects like mosaic are is searched with the "@" operator. These attribute names are analogous to the columns of a dataframe in a relational database structure.

Once downloaded, you should be able to type statements mosaicdatabase@[insertfield] (where [insertfield] is a particular trait). If you are working in Rstudio, after the "@" a drop-down of the slots (traits) should autopopulate.

```{r eval=FALSE}
# Three examples of queerying traits (easiest for navigation)
mosaic@biomass
mosaic@height
mosaic@volancy
```

You can also load the data using slots. As a matter of best-practice, slot inquiry is the suggested practice -- it has the disadvantage of not offering an autopopulation of the attributes contained in the database (traits must be spelled out manually).

```{r eval=FALSE}
# You can also search these by slot (recommended)
slot(mosaic, "biomass")
slot(mosaic, "height")
slot(mosaic, "volancy")
```

Within each trait object, there are eight fields. The first of these is called "values" and contains the data.The other seven - "value", "author", "year", "journal, "doi", "database", and "mosaic - are metadata corresponding with each value record. So, for instance: 

```{r eval=FALSE}
mosaic@species[[2]] # For this species, let us look at volancy (flight capacity) value
mosaic@volancy@value[[2]] # or, equivalently:
slot(slot(mosaic, "volancy"), "value")[[2]]
```
Corresponds with the following metadata

```{r eval=FALSE}
mosaic@volancy@author[[2]] # for "Acinonyx jubatus"
slot(slot(mosaic, "volancy"), "author")[[2]] # Author of the source publication
slot(slot(mosaic, "volancy"), "year")[[2]] # And year of the source publication
slot(slot(mosaic, "volancy"), "journal")[[2]] # The journal of the source publication
#etc.
```


: (1) values (which record the records); (2) metadata (which records the source/provinance of the data). Values are unitless values (either numeric or factorial) that are reported in units described in the **Mosaic User Guide** <http://mosaicdatabase.web.ox.ac.uk/user-guide>. The metadata is organised into additional attributes, reflecting the individual elements of the metadata for a given record, including the authors, journal, year of publication, databases from which data are sourced (if applicable)


```{r eval=FALSE}
# Three examples of queerying traits
mosaic@biomass@values
mosaic@biomass@authors
mosaic@biomass@year
mosaic@biomass@journal
```



```{r eval=FALSE}
# Let us look at a particular record for height
mosaic@height@values[[5]]
mosaic@height@authors[[5]]

# This also works with the slot format
slot(mosaic, "height"), "journal")[[5]]
slot(mosaic, "height"), "database")[[5]]
```


## Using MOSAIC functions to quickly access files

A series of convenience functions are scripted to facilitate navigating and working with the mosaic database that can be accessed by running the follwing script.

Below we highlight some of the basic queries for which the mosaic functions can assist.
```{r include=FALSE}
spp_check <- function(binomialName){ # is any given species in the database
  return(length(which(mosaic@species==binomialName))>0)
}

traitAllRecords <- function(trait){
  return(slot(slot(mosaic, trait),"value"))
}

speciesSummary  <- function(speciesNameLatin){
  speciesIndex <- which(mosaic@species==speciesNameLatin)
  metadataNames <- slotNames(mosaic)[-c(1:3)]
  meta <- data.frame(matrix(NA, nrow=1, ncol=length(metadataNames)))
  colnames(meta) <- metadataNames
  for(i in 1:length(metadataNames)){
    meta[,i] <- slot(slot(mosaic, metadataNames[[i]]), "value")[[speciesIndex]]
  }
  return(meta)
}

multiSummary <- function(sppList){
  summary <- lapply(sppList, speciesSummary)
  framesummary <- data.frame(matrix(NA, nrow=length(summary), ncol=length(summary[[1]])+1))
  colnames(framesummary) <- c("sppnames", colnames(summary[[1]]))
  for(i in 1:length(summary)){
    framesummary[i,] <- c(sppList[[i]],summary[[i]])
  }
  return(framesummary)
}


TraitFrequency <- function(trait){
  focalTrait <- slot(slot(mosaic, trait), "value")
  focalTrait <- as.factor(focalTrait)
  hold <- matrix(NA,
                 nrow=length(levels(focalTrait)),
                 ncol=2)
  
  rownames(hold) <- levels(focalTrait)
  colnames(hold) <- c("counts", "freq")
  
  for(i in 1:length(levels(focalTrait))){
    hold[i,1] <- sum(focalTrait==levels(focalTrait)[[i]])
  }
  
  hold[which(rownames(hold)=="NDY"),2] <- NA
  hold[-which(rownames(hold)=="NDY"),2] <- round(hold[-which(rownames(hold)=="NDY"),1]/sum(hold[-which(rownames(hold)=="NDY"),1]), 3)
  return(hold)
}

metaRecords <- function(trait, index){
  metas <- data.frame(matrix(NA, nrow=1, ncol=5))
  colnames(metas) <- c("author", "year", "journal", "database", "mosaic")
  metas[1,1] <- slot(slot(mosaic, trait), "author")[[index]]
  metas[1,2] <- slot(slot(mosaic, trait), "year")[[index]]
  metas[1,3] <- slot(slot(mosaic, trait), "journal")[[index]]
  metas[1,4] <- slot(slot(mosaic, trait), "database")[[index]]
  metas[1,5] <- slot(slot(mosaic, trait), "mosaic")[[index]]
  return(metas)
}

multiMetaRecords <- function(trait, sequenceOfIds){
  formerMeta <- metaRecords("volancy", sequenceOfIds[[1]])
  for(i in 2:length(sequenceOfIds)){
    formerMeta <- rbind(metaRecords("volancy", sequenceOfIds[[i]]), formerMeta)
  }
  return(formerMeta)
}
```

### Is a species included in Mosaic?
```{r}
spp_check("Fritillaria biflora")
spp_check("Ursus meritimus")
```

### Can I see all records for a given trait?
```{r eval=FALSE}
traitAllRecords("biomass") # Resulting values not shown for space
```
```{r echo=FALSE}
head(traitAllRecords("biomass")) # Resulting values not shown for space
```

### Can I see an overview of all records for a given species?
```{r}
speciesSummary("Aepyceros melampus")
```

### Can I see all records for more than one species?
```{r}
multiSummary(c("Acinonyx jubatus", # you can also pass lists or dataframes to this command
               "Acropora downingi",
               "Aepyceros melampus",
               "Alces alces",
               "Alligator mississippiensis"))
```

### Can I get a breakdown of counts/frequency of trait values?
```{r}
TraitFrequency("volancy")
TraitFrequency("growthdet")
TraitFrequency("hermaph")
```

### Can I get all metadata for one or more traits?
```{r}
metaRecords("volancy", 14)
multiMetaRecords("volancy", c(14:20))
```


## Integrating MOSAIC with Structured Population Models
MOSAIC is motivated by the need for trait integration with existing population databases, the COMADRE-COMPADRE-PADRINO databases in particular. Please carefully read the user guide and information at <https://compadre-db.org/> and <https://padrinodb.github.io/Padrino/> before working with COMADRE and COMPADRE databases.

MOSAIC is organised as a relational database that is linked with COMADRE and COMPADRE through species name and unique-identifiers of the matrices.


```{r eval=FALSE}
link_comadre(21)
```
